---
title: "Census Data Analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Function to install packages if they are not already installed
install_if_missing <- function(p) {
  if (!requireNamespace(p, quietly = TRUE)) {
    install.packages(p, dependencies = TRUE)
  }
}

# List of required packages
packages <- c("ggplot2", "rmarkdown", "reshape2")

# Install missing packages
invisible(lapply(packages, install_if_missing))

# Load the libraries
library(ggplot2)
library(rmarkdown)
library(reshape2)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Load and Clean Data

Load and Clean Data can be done as follows:

```{r}
dataset1 <- read.csv("../data/census_2021_2001.csv", TRUE, sep = ",")
finaldata <- na.omit(dataset1)
dataset2021 <- subset(finaldata, year != "2001")

```

## Fertility Rate Frequency Distribution

You can also embed plots, for example:

```{r}
fertilityRateHist <- hist(dataset2021$total.fertility.rate, 
  breaks = "Scott", freq = NULL, 
  include.lowest = TRUE, right = TRUE, 
  main = "Fertility Rate Frequency", 
  xlab = "Fertility Rate", 
  ylab = "Frequency", 
  col = "#999999"
)

abline(v = mean(dataset2021$total.fertility.rate), col = "red", lwd = 2)
abline(v = median(dataset2021$total.fertility.rate), col = "royalblue", lwd = 2)
legend(x = "topright", c("Mean", "Median"), col = c("red", "royalblue"),
 lwd = c(2, 2))

```

## Life Expectancy Frequency Distribution

You can also embed plots, for example:

```{r}
expectancyAllHist <- hist(dataset2021$life.expectancy.both.sexes, 
  breaks = "Scott", freq = NULL, 
  include.lowest = TRUE, right = TRUE, 
  main = "Life Expectancy Frequency Both Sexes", 
  xlab = "Life Expectancy Rate", 
  ylab = "Frequency", 
  col = "#999999"
)

abline(v = mean(dataset2021$life.expectancy.both.sexes), col = "red", lwd = 2)
abline(v = median(dataset2021$life.expectancy.both.sexes), col = "royalblue", lwd = 2)
legend(x = "topright", c("Mean", "Median"),
 col = c("red", "royalblue"), lwd = c(2, 2))

```

## Statistical Analysis
```{r}
cat("Total Fertility Rate:\n")
print(mean(dataset2021$total.fertility.rate))
print(sd(dataset2021$total.fertility.rate))
print(quantile(dataset2021$total.fertility.rate))

cat("Life Expectancy Both Sexes:\n")
print(mean(dataset2021$life.expectancy.both.sexes))
print(sd(dataset2021$life.expectancy.both.sexes))
print(quantile(dataset2021$life.expectancy.both.sexes))

cat("Life Expectancy Males:\n")
print(mean(dataset2021$life.expectancy.males))
print(sd(dataset2021$life.expectancy.males))
print(quantile(dataset2021$life.expectancy.males))

cat("Life Expectancy Females:\n")
print(mean(dataset2021$life.expectancy.females))
print(sd(dataset2021$life.expectancy.females))
print(quantile(dataset2021$life.expectancy.females))


```

## Correlation Analysis

```{r}
ggplot(data = dataset2021, aes(x = total.fertility.rate, y = life.expectancy.both.sexes)) +
  geom_point(size = 2, alpha = .75) +
  ggtitle(paste("Correlation:", round(cor(dataset2021$total.fertility.rate, dataset2021$life.expectancy.both.sexes), digits = 2))) +
  xlab("Total Fertility Rate") +
  ylab("Life Expectancy of Both Sexes")


```

## Grouped Barplots

```{r}
# Base R approach to achieve similar results as tidyr and dplyr

# Fertility rate in each subregion
boxplot(dataset2021$total.fertility.rate ~ dataset2021$subregion,
        main = "Fertility rate in each subregion",
        xlab = "Subregion",
        ylab = "Fertility Rate",
        las = 2, # Make labels perpendicular to the axis
        col = "lightblue")

# Life expectancy in each subregion
melted_data <- melt(dataset2021, id.vars = c("country", "genc", "subregion", "region", "year"),
                    measure.vars = c("life.expectancy.males", "life.expectancy.females"))
boxplot(value ~ subregion + variable, data = melted_data,
        main = "Life expectancy in each subregion",
        xlab = "Subregion",
        ylab = "Life Expectancy",
        las = 2, # Make labels perpendicular to the axis
        col = "lightblue")

```